Fail Faster

### 2025-03-21 paper/paper.tex
intro

Property-based testing (PBT) is a software testing technique that uses
random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} in industry have shown that
practitioners often interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This means that PBT is often given a very short time
window to find bugs --- e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}

\bcp{Feels like we're missing a trick in the transition here.  Can we
  say, ``Because performance is important, PBT libraries are already
  quite tuned.  [Can we substantiate that??]  But more is
  possible...'' }

One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
that provide combinators for building and composing generators. Optimizing these
libraries is a key opportunity to speed up PBT: if programmers are
provided with the tools to write fast generators, then they do not themselves
need to be experts in performance tuning to find bugs fast.

However, careful measurements show that the performance of existing generator libraries falls significantly short of what
is possible. This is due to two major sources of inefficiency.
First, the same high-level design approach that gives generator libraries their
flexibility also introduces layered abstractions and indirections that are difficult for
compilers to optimize, shifting work to run time.
Second, the \rand that the generator uses to make choices can
constitute an unexpectedly high proportion of its runtime.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi
Together, these observations raise a fundamental question: can generator libraries be expressive and high-level without compromising efficiency?

To address this question, we outline two key principles for designing
generator libraries that marry efficiency with flexibility.
First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used to
build DSLs without abstraction overhead (``without regret''~\cn{}).
We apply this technique to generators, completely eliminating the run-time overhead of
many common generator abstractions.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison, showing that showing that certain 
implementations significantly reduce overhead.
Together, these insights highlight practical strategies for improving generator libraries, 
combining the efficiency gains of staging with a careful evaluation of \rand to optimize performance.

To demonstrate the effectiveness of our approach, we build two different PBT libraries in two different langauges.
In OCaml, we build \camlname{}, based on \bq{}  --- a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
In Scala 3, we build \scalaname{}, based on ScalaCheck --- a very
mature PBT library and the {\em de facto} choice\bcp{the wording here
  implies that there are other choices in Scala, some of which might
  be reasonable, and tht base\_quickcheck is not the de facto choice
  in OCaml...} in Scala \cn{}.
In both languages, our libraries are direct drop-in replacements that are
\textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with our technique
produces exactly the same sequence of values as the same generator written with the original generator library.
The semantic equivalence between the original generators and \name{} generators allows us to
perform clean, fine-grained comparisons, as differences in generation speed are
solely attributable to the staging and \rand optimizations.

>>>>

Property-based testing (PBT) is a software testing technique that uses
random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} in industry have shown that
practitioners often interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This means that PBT is often given a very short time
window to find bugs---e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
The faster a test fails, the better.
% The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}
\iflater 
\bcp{Feels like we're missing a trick in the transition here.  Can we
  say, ``Because performance is important, PBT libraries are already
  quite tuned.  [Can we substantiate that??]  But more is
  possible...'' }
\fi

One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
that provide combinators for building and composing generators. However, careful measurements show that the performance of existing generator libraries falls far short of what
is possible.

This is due to two major sources of inefficiency.
First, the same high-level design approach that gives generator libraries their
flexibility also introduces layered abstractions and indirections that are difficult for
compilers to optimize, shifting work to run time.
Second, the \rand that a generator uses to make choices can
constitute an unexpectedly high proportion of its runtime. Combined, these factors limit 
the effectiveness of PBT.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi

We lay a path forward.
First, we show how to use \textit{multi-stage programming}---a classic technique for 
building fast DSLs “without regret”~\cn{}---to eliminate the overhead of 
many common generator abstractions.
Second, we precisely characterize the impact of \rand choice through a natural experiment that preserves 
generator behavior while improving execution speed, demonstrating that the entropy source itself should be treated as a first-class 
concern in generator library design.
Taken together, these insights can be applied to create tomorrow's generator libraries: those that 
combine high-level design with low-level performance.

To demonstrate the effectiveness and generality of our approach, which we call \name, we implement staged PBT libraries in multiple languages.
In OCaml, we build \camlname{}, based on \bq{}---a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
In Scala 3, we build \scalaname{}, based on ScalaCheck---a very
mature PBT library and the {\em de facto} choice\bcp{the wording here
  implies that there are other choices in Scala, some of which might
  be reasonable, and tht base\_quickcheck is not the de facto choice
  in OCaml...} in Scala \cn{}.
In both languages, our libraries are direct drop-in replacements that are
\textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with our technique
produces exactly the same sequence of values as the same generator written with the original generator library.
The semantic equivalence between the original generators and \name{} generators allows us to
perform clean, fine-grained comparisons, as differences in generation speed are
solely attributable to the staging and \rand optimizations.
